---
title: "The decline of full-time employment"
date: 4/07/2024
editor: visual
knitr:
  opts_chunk: 
    dev: "ragg_png"
execute: 
  echo: false
fig-width: 6
fig-asp: 0.618
fig-align: center
---

```{r}
#| label: setup

library(reportabs)
library(tidyverse)
library(aititheme)
library(strayr)

labour_force <- read_absdata("labour_force", export_dir = "data")
hours_worked <- read_absdata("hours_worked", export_dir = "data")
industry_employment <- read_absdata("industry_employment", export_dir = "data")


```

The proportion of workers employed full-time has declined significantly in Australia, even over the past two decades (see @fig-full-time-proportion). Where full-time employment once comprised more than 80% of total employment, it is today below 69% . Increased economic activity and labour demand in the post pandemic period resulted in a temporary increase in full-time employment, to over 70%, a level not seen since 2010, but these gains appear to have been short lived.

```{r}
#| label: fig-full-time-proportion
#| fig-cap: Proportion of full-time employment to total employment, Australia, Trend

labour_force |> 
  filter(indicator %in% c("Employed full-time", "Employed part-time"),
         state == "Australia", 
         series_type == "Trend", 
         age == "Total (age)",
         sex == "Persons") |> 
  pivot_wider(names_from = indicator, 
              id_cols = c(-series_id, - series)) |> 
  mutate(part_time_share_total = `Employed full-time`/(`Employed full-time` + `Employed part-time`)) |> 
  pivot_longer(cols = "Employed full-time":"part_time_share_total",
               names_to = 'indicator',
               values_to = "value") |> 
  mutate(unit = ifelse(indicator == "part_time_share_total", "Percent", unit)) |> 
  abs_plot(v = list(indicator = "part_time_share_total"),
           years = 1999,
           colour = "Concrete Grey",
           legacy = TRUE) +
  scale_y_continuous(labels = scales::percent_format())
```

While total employment has continued to increase, this has mostly been driven by increases in part-time employment.

-   In January, part-time employment in Australia was `r as_comma(1000*value_at(labour_force, list(indicator = "Employed part-time", state = "Australia"), at_year = 2024, at_month = "January"))`, increasing to `r as_comma(1000*value_at(labour_force, list(indicator = "Employed part-time", state = "Australia"), at_year = 2024, at_month = "May"))` by May.

-   At the same time, full-time employment increased from `r as_comma(1000*value_at(labour_force, list(indicator = "Employed full-time", state = "Australia"), at_year = 2024, at_month = "January"))` to `r as_comma(1000*value_at(labour_force, list(indicator = "Employed full-time", state = "Australia"), at_year = 2024, at_month = "May"))`.

```{r}
#| label: fig-pt-ft-employment-sa 
#| fig-cap: Part-time and full-time employment in Australia 
#| out-width: "70%"  
 
abs_plot(labour_force, 
         years = 2020,
         v = list(indicator = c("Employed full-time", "Employed part-time"),                    state = "Australia"),          
         compare_aus = FALSE, 
         colour = "Concrete Grey",
         legacy = TRUE) +  
  theme(legend.position = 'bottom')    
```

The growth of part-time employment at the expense of full time employment may simply reflect individuals preferences to work part-time. However, if full-time work is desired but not available, this may reveal increasing insecurity in work.

```{r}
#| label: fig-unemployment-rate
#| fig-cap: Unemployment rate, Australia, Trend

abs_plot(labour_force,
         legacy = TRUE,
         v = list(indicator = c("Unemployment rate looked for full-time work", "Unemployment rate looked for only part-time work"),
                  series_type = "Trend",
                  state = "Australia")) +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(nrow = 2))
```

## Industry

```{r}
#| label: fig-aus-industry
#| fig-cap: Proportion of full-time employment to total employment, Australia, by industry

# five largest industries employing in SA
industry_large <- industry_employment |> 
  filter(state == "Australia",
         industry != "Total (industry)",
         indicator == "Employed total",
         date == max(date)) |> 
  slice_max(value, n = 5) |>  
  pull(industry)

industry_employment |> 
  filter(state == "Australia", 
         industry != "Total (industry)", 
         indicator != "Employed total")  |> 
  inner_join(distinct(anzsic2006, anzsic_division, anzsic_division_code), 
             by = c("industry" = "anzsic_division")) |> 
  arrange(anzsic_division_code) |> 
  mutate(industry = fct_inorder(industry))  |> 
  pivot_wider(names_from = indicator,
              values_from = value,
              id_cols = c(date, state, series_type, industry, unit)) |> 
  mutate(part_time_share_total = `Employed full-time`/(`Employed full-time` + `Employed part-time`)) |> 
  pivot_longer(cols = "Employed full-time":"part_time_share_total",
               names_to = "indicator") |>
  filter(industry %in% industry_large) |> 
  abs_plot(v = list(indicator = c("Employed full-time", "Employed part-time"),
                    state = "Australia"),
           years = 2020,
           compare_aus = F,
           legacy = TRUE) +
  aes(col = indicator) 
```

## Unemployment

```{r}

```

```{r}
#| label: hours-worked-per-person


hw <- hours_worked |> 
  filter(state == "South Australia",
         sex == "Persons", 
         series_type == "Trend") |> 
  pivot_wider(id_cols = c(date, sex, state, series_type),
              values_from = value,
              names_from = indicator)

lf <- labour_force |> 
  filter(state == "South Australia",
         indicator == "Employed total",
         series_type == "Trend",
         sex == "Persons") |> 
  pivot_wider(id_cols = c(date, sex, state, series_type),
              values_from = value,
              names_from = indicator)

hours_worked_per_person <- inner_join(hw, lf)


```

## Sahm Recession Indicator

```{r}
#| label: fig-sahm-sa
#| fig-cap: Sahm recession indicator, South Australia
#| out-width: "70%"

sahm("South Australia")

```
