---
title: "Made in Australia"
author: "Hamish Gamble"
date: "2024-06-26"
categories: [analysis]
fig-width: 4
fig-asp: 0.618
---

```{r}
#| label: setup 
#| 
library(ecomplexity)
library(readabs)
library(tidyverse)
library(ggplot2)
library(reportabs)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

emp_ind <- read_abs("6291.0.55.001", 
                    tables = "4", 
                    path = "data", 
                    check_local = TRUE)  |> 
  separate_series(column_names = c("industry_division", "indicator")) |> 
  mutate(industry_division = case_when(
    is.na(industry_division) ~ "Total (industry)",
    TRUE ~ industry_division
  )) 
```

The Future Made in Australia agenda is a \$22.7 billion policy aimed to increase local participation in the net zero supply chain, and develop domestic capabilities in industries which can provide economic resilience and security.

A country with low manufacturing capabilities, and low economic complexity like Australia, requires a directed and deliberate industry policy to create an economy that is more resilient, and more capable of addressing global challenges such as decarbonisation.

## Manufacturing

Decline in manufacturing employment

```{r}
#| label: fig-manufacturing-emp
#| fig-cap: Manufacturing employment, Trend, Australia
#| out-width: "70%"
#| fig-align: 'center'


emp_ind |>
  filter(series_type == "Trend", 
         industry_division == "Manufacturing")  |>
  ggplot(aes(x = date, y = value, col = industry_division)) +
  labs(title = "Manufacturing Employment") +
  geom_line() +
  theme_fof() + 
  scale_y_continuous(labels = scales::comma_format(suffix = "k"),
                     name = NULL) + 
  scale_x_date(name = NULL,
               labels = scales::date_format("%b-%Y")) +
  scale_colour_fof()
```

```{r}
#| label: fig-manufacturing-share
#| fig-cap: Manufacturing employment, share of total employment, Trend, Australia
#| out-width: "70%"
#| fig-align: "center"

emp_ind |> 
  filter(industry_division != "Total (industry)") |> 
  group_by(series_type, date) |>
  mutate(share = value / sum(value)) |> ungroup() |> 
  filter(industry_division == "Manufacturing", 
         indicator == "Employed total",
         series_type == "Trend") |> 
  ggplot(aes(x = date, y = share, col = industry_division)) + 
  geom_line() + 
  theme_fof() +
  labs(title = "Manufacturing Share of Employment") +
  scale_y_continuous(labels = scales::percent_format(),
                     name = NULL) + 
  scale_x_date(name = NULL,
               labels = scales::date_format("%b-%Y")) + 
  scale_colour_fof()

```
